---
title: "Computation Workshop 5 - Quick R"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

This is a notebook containing:

1.  Basic R commands for beginners
2.  Phylogeny visualization

The notebook displays both the commands and their results automatically.
For best practice, feel free to modify the examples and experiment with
the code throughout the tutorial.

# Basic Commands

### What is R & RStudio?

R is a free programming software generally used for data analysis and
creating graphics.

RStudio is a user-friendly interface for R.

On RStudio you see multiple panels:

-   **Console Pane:** any running command, errors, outputs can be seen
    here

-   **Environment/Workspace/History Pane:** the defined variables are
    stored here. You can check their dimensions, length, size etc. You
    can also click on matrix or data frame to display them on source
    pane in more detail.

-   **File/Plot/Package/Help/Viewer Pane:**

    -   Files tab shows the structure of current working folder

    -   Plot tab preview the graphs; Packages to check what packages
        were installed and loaded

    -   Help to search any function and check how it is used (generally
        there are examples at the end of each help pages)

-   **Source Pane:** (you will see if you click on hidden panel or open
    "New R Script"); you can write your commands here in script,
    markdown or notebook form.

    <div>

    > Tip: It is better to write your commands on R script file on
    > source pane rather then just running them on the console. With
    > script every action will be stored in a file for further usage or
    > sharing. It is important to save the script file regularly when
    > changes are made.

    </div>

### R Basics

R can do any mathematical calculations and print any text in double
quotes.

We sometimes use comments to explain what R code does. Some examples
below:

```{r}

1020 * 1093
108420 / 2
240 + 1234
1197- 129

"Hello World" # This is a comment, R won't run this

# This is also a comment

```

It gives an error when numbers are placed in quotation marks, because
they are then treated as characters rather than numeric values.

```{r}

"1020"*"1093"

```

It also gives an error when non-variable words are written without
quotation marks.

```{r}

Hello
```

### Variables

Variables stores the data values. If you assign a value to any
word/letter, it becomes a variable.

```{r}
name
```

Opps, error. Let's try again.

```{r}
name = "Evrim" 

name

print(name)
```

> Tip: The more descriptive the variable, the better. Instead of using
> letters (x=, y= etc.) , it is better to use words (student_name,
> student_age etc.). Names should start with letter but can be a
> combination of letters and digits. Variable names are case sensitive
> (i.e., name, Name, NAME are different variables). The names of R
> functions cannot be used as variable name.

### Data Types

We can check the data type using class() function

```{r}
x=1.5
class(x)

x="workshop"
class(x)

x=TRUE
class(x)
```

### If / Else if / Else statements

Logical values are important to define conditions (i.e., if - else
statements). A condition is evaluated whether it is true or false, "IF"
it is true, the next block of code will be executed.

```{r}

a=40
b=100

if(b>a){
  print("b is greater than a")
} else if (a == b) {
  print("a and b are equal")
} else {
  print("a is greater than b")
}
```

We can combine multiple conditions with AND and OR logical operators.

```{r}
x=50
y=20
z=-40

if(x>y & z<y){
  print("TRUE")
}

if(x>y | z>y){
  print("TRUE")
}
```

### For Loop and Functions

We can iterate the same action for a sequence of elements using for
loop.

```{r}

fruits = list("apple", "banana", "strawberry")

for (item in fruits){
  print(item)
}
```

We can define our own function with a block of code that can be used
with a given parameter (argument).

```{r}

some_math=function(argument1, argument2){
  result=((argument1-argument2)/(argument1+argument2))*100
  return(result)
}

some_math(20,10)
some_math(100,12103)
```

### Data Structures

R has different data structures to handle and store the values. These
are vectors, lists, matrices, arrays and data frames.

-   **Vector\*:** list of items of the same type

-   **List:** can combine different types of data

-   **Matrix\*:** 2D data structure where all elements are the same time

-   **Array:** similar to matrix but have more dimensions, stores
    element of the same type in multiple dimensions

-   **Data frame\*:** similar to matrix but it can hold different types
    of data in different columns

\* - mostly used data structures

```{r}

# vector of strings

colors=c("red","blue","green")
colors

# vector of numbers
some_numbers=c(11,23,34,45,56)

```

```{r}
1:5

seq(1:5)

seq(from=2,to=10,by=2)

rep(1,5)

rep(2,5)

c(rep(1,5), rep(2,5))

rep(c(1,2))

rep(c(1,2), each=5)

```

```{r}
# each element has a position in the vector = index
some_numbers[1] 
some_numbers[3]
some_numbers[-2] # removed the second element
some_numbers[2:5] # prints the items between second and the fifth elements
```

```{r}
# list of different items
alist=list(c("apple","banana"),"color",c(1,2))
alist

```

```{r}
alist[[1]]
alist[[2]]
alist[[3]][1]

```

```{r}
mat=matrix(c(1,2,3,4,5,6), nrow=3, ncol=2)
mat

mat=matrix(c(1,2,3,4,5,6), nrow=3, ncol=2, byrow=TRUE)
mat

```

```{r}
# dimensions of the matrix (row numbers, column numbers)
dim(mat)
# number of rows
nrow(mat)
# number of columns
ncol(mat)

# the item in first row, second column
mat[1,2]

# all items in the first column
mat[,1]

# all items in the third row
mat[3,]
```

```{r}
# An array with more than one dimension
multiarray <- array(c(1:24), dim = c(4, 3, 2))
multiarray

multiarray[2,3,2]
```

```{r}
data=data.frame(
  Training=c("strength","stamina","other"),
  Pulse=c(c(100,50,120)),
  Duration=c(60,30,45)
)
data

dim(data)
data[1,2]
data$Training



```

### Install Packages

```{r}
install.packages("ggtree")
install.packages("seqinr")

```

### Read and Save Data

```{r}
data=read.table("/Users/fer/Desktop/week4/IF2_NTD_42C_cleaned.csv")
read.csv("/Users/fer/Desktop/week4/IF2_NTD_42C_cleaned.csv")
library(seqinr)
read.fasta("/Users/fer/Desktop/week4/")
library(ggtree)
read.tree("/Users/fer/Desktop/week4/Rca_phy.treefile")
```

# Introduction to generating phylogenetic Trees in R

```{r}
options(
  repr.plot.width  = 10,   # increase this for wider plots
  repr.plot.height = 8,    # increase this for taller plots
  repr.plot.res    = 150   # bump resolution (DPI); 150–200 is usually good
)
```

### Setup: Installing and loading packages

```{r}

# Install BiocManager if needed
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# CRAN packages
install.packages(c(
  "ggplot2", "tidyr", "dplyr", "ape",
  "ggnewscale", "scales"
))

# Bioconductor: ggtree
BiocManager::install("ggtree")

# Load libraries
library(ggtree)
library(ape)
library(ggnewscale)
library(dplyr)
library(tidyr)
library(scales)
library(ggplot2)

```

### Reading the Phylogenetic Tree

```{r}
phy <- read.tree(file = "/Users/fer/Desktop/week4/Rca_phy.treefile")

phy
```

### Read the metadata table as a dataframe

```{r}
df = data.frame(read.delim("/Users/fer/Desktop/week4/Rca_info.tsv"))

head(df)
```

```{r}
annot_df <- df %>% select(MSA_id, slen, major_group, domain, kingdom, phylum) %>% rename(label = MSA_id)

head(annot_df)
```

### Plotting the Phylogenetic Tree

```{r}
phy_rect <- ggtree(phy)

phy_rect
```

### Adding other tip labels and scale to the phylogeny

```{r}
phy_rect_tips <- ggtree(phy) +
  geom_treescale() +    ## Adds the scale
  geom_tiplab(size = 2) ## Adds the tip labels

phy_rect_tips
```

### Plot it in a circular layout

```{r}
phy_circ <- ggtree(phy, layout = "circular")

phy_circ
```

### Associating the metadata with the phylogenetic tree

```{r}
p <- ggtree(phy) %<+% annot_df + geom_treescale()

p
```

### Coloring the tips by categorical data

```{r}
p_domain <- p + geom_tippoint(aes(color=domain), size = 3)

p_domain
```

```{r}
p_kingdom <- p + geom_tippoint(aes(color=kingdom), size = 3)

p_kingdom
```

```{r}
p_circ <- ggtree(phy, layout = "circular") %<+% annot_df + geom_treescale()

p_kingdom_circ <- p_circ + geom_tippoint(aes(color=kingdom), size = 3)

p_kingdom_circ
```

### Mapping continuous data with the phylogeny

-   Coloring the tips based on sequence length

```{r}
p_slen_color <- p +
  geom_tippoint(aes(color = slen), size = 3) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(color = "Sequence length") +
  theme(legend.position = "right")

p_slen_color
```

-   Tip sizes by sequence length

```{r}
p_slen_size <- p +
  geom_tippoint(aes(size = slen), alpha = 0.7) +
  scale_size_continuous(range = c(1, 5)) +
  labs(size = "Sequence length") +
  theme(legend.position = "right")

p_slen_size
```

### Adding a heatmap with our phylogeny

```{r}
## Creating a minimal df only with the slen information

heat_df <- annot_df %>%
  select(label, slen) %>%
  distinct(label, .keep_all = TRUE) # Ensure unique labels before setting as row names

rownames(heat_df) <- heat_df$label
# row.names(heat_df) <- heat_df$label
heat_df$label <- NULL

head(heat_df)
```

```{r}
p_base <- ggtree(phy) %<+% annot_df + geom_treescale()

# Add heatmap of sequence length
p_heat <- gheatmap(p_base, heat_df,
                   offset = 0.05, width = 0.1,
                   colnames = TRUE) +
  scale_fill_gradient(low = "lightyellow", high = "red",
                      name = "Seq length")

p_heat
```

### Combining the Heatmap and the colored tips

```{r}
p_kingdom <- p + geom_tippoint(aes(color=kingdom), size = 3)

p_kingdom_heatmap <- gheatmap(p_kingdom, heat_df,
                   offset = 0.05, width = 0.1,
                   colnames = TRUE) +
  scale_fill_gradient(low = "lightyellow", high = "red",
                      name = "Seq length")

p_kingdom_heatmap
```

### Saving the plots

```{r}
ggsave("/content/phylogenetic_tree_with_kingdom_and_heatmap.png", plot = p_kingdom_heatmap, width = 10, height = 8, dpi = 300)
ggsave("/content/phylogenetic_tree_with_kingdom_and_heatmap.svg", plot = p_kingdom_heatmap, width = 10, height = 8, dpi = 300)
```
